|

### 2021-08-10 prova_knit_finale.Rmd
change

A second approach can be used when sampling has to be done sequentially: \textit{sequential importance resampling} (SIR). Differently from SIS, an intermediate resampling stage takes place, in which time 0 values are sampled with replacement according to a given resampling strategy (usually, sampling indexes from a multinomial distribution with probabilities equal to the weights computed through IS at time 0).\
While in the SIS all the particles (i.e., the sampled values) "survive", with the "risk" that particles associated with small weights are carried-over (in the sense that sampling in period 1 occurs from the kernel $M_1(X_0^n,dx_1)$), the same does not occur in SIR. In this second algorithm, particles with larger weights are more likely to be resampled and be used to ``generate" the particles in the next period.\
We describe the SIR more formally, focusing on the previous two-period case.
\begin{enumerate}
    \item Sample $N$ values $X_0$ from the proposal measure $\M_0$: $X_0^n\thicksim \M_0$ for $n=1,...,N$.
    \item Compute the weights $W_0^n$ for $n=1,...,N$.
    \item Sample $N$ indexes $(A_1^1,...A_1^N)$ from a multinomial distribution such that for all $j=1,...N$, $Pr(\{A_1^j=n\})=W_0^n$, for all $n=1,...N$.
    \item Build a new time 0 sample, using the values for $X_0$ associated to the previously extracted indexes $(A_1^n)_{n}$: $(X_0^{A_1^1},...,X_0^{A_1^N})$.
    \item Sample $N$ values for $X_1$ from the transition kernel $M_1$: $X_1^n\thicksim M_1(X_0^{A_1^n},dx_1)$ for $n=1,...,N$.
\end{enumerate}
Steps 3-4 constitute the actual resampling ones. At this point, let us define the new two-period sequence $\tilde X_{0:1}^n:=(X_0^{A_1^n},X_1^n)$. Then, the particle approximation for the joint distribution of $(X_0,X_1)$ becomes:
\begin{equation*}
     \Q_{1}^N(dx_{0:1})=\frac{1}{N}\sum_{n=1}^N\delta_{\tilde X_{0:1}^n}(dx_{0:1})
\end{equation*}
Note that, after resampling occurred, each particle $\tilde X_{0:1}^n:=(X_0^{A_1^n},X_1^n)$ has the same probability weight.\
To understand the reason of the previous statement, let us assume that both $\Q_0$ and $\M_0$ admit densities $q_0$, $m_0$ with respect to the same dominating measure $\nu(dx_0)$ and that the transition kernel $M_1$ also admits the density $m_1$. 
Consider the sequential updating for the weights:
\begin{equation}
    w^n_1=\frac{\Q_1(dx_{0:1})}{\M_1(dx_{0:1})}\propto \frac{q_0(x_0^{A_1^n})}{Pr(A_1^n)\cdot m_0(x_0^{A_1^n})}\cdot \frac{m_1(x_1^n|x_0^{A_1^n})}{m_1(x_1^n|x_0^{A_1^n})}
=\underbrace{\frac{q_0(x_0^{A_1^n})}{ m_0(x_0^{A_1^n})}}_{W_0^n}\cdot \underbrace{\frac{1}{Pr(A_1^n)}}_{1/W_0^n}=1\label{sis_update}
\end{equation}
where $\M_1$ denotes the probability measure for the extraction of the particle $\tilde x_{0:1}^n$ according to SIR. $\M_1$ (second equality sign in the previous formula) can be rewritten as the product between the kernel for $x_1^n$ given $x_0^{A_1^n}$ and the probability of $x_0^{A_1^n}$ itself. This last probability can also be rewritten as the product between the proposal density $m_0$ from which $X_0$ is extracted and the probability that it is sampled again when resampling occurs.

\subsection{SIS vs SIR}
The choice between the two sequential sampling strategies (SIS and SIR) has significant implications in terms of computational cost and variance of the importance weights. In fact, it is clear that SIR is more costly computationally since it requires an intermediate resampling step. Nonetheless, SIS suffers from the problem knonw as \textit{curse of dimensionality}, i.e., the fact that the $ESS$ collapses over time as the variance of the weights diverges.\
Consider an extension of SIS to a $T$-period model and let us assume that the target and the proposal kernels admit densities, respectively, $q_t$ and $m_t$ with respect to the same dominating measures (this implies that the joint distributions\footnote{Let us denote the joint distributions with $q_t$ and $m_t$ too, with some abuse of notation.} admit a density too). Then, weights, for every $t=1,...,T$, are updated according to:
\begin{equation}
    w_t^n\propto \frac{q_t(x_{0:t}^n)}{m_t(x_{0:t}^n)}\propto \underbrace{\frac{q_{t-1}( x_{0:t-1}^n)}{m_{t-1}(x_{0:t-1}^n)}}_{\propto w_{t-1}^n}\cdot \frac{q_t( x_t|x_{t-1}^n)}{m_t(x_t|x_{t-1}^n)}\propto 
    w_{t-1}^n\cdot\frac{q_t( x_t|x_{t-1}^n)}{m_t(x_t|x_{t-1}^n)}\label{sis_update}
\end{equation}
Equation \eqref{sis_update} suggests, intuitively, that over time new variance is added from the incremental weights (i.e., the updating factor $q_t( x_t|x_{t-1}^n)/m_t( x_t|x_{t-1}^n)$) determining the curse of dimensionality issue. Instead, when resampling takes place, since the past weights $w_{t-1}$ are reset to be equal across them all the variance of $w_t$ inherited from the past is shut-down and the only source of variability at each stage comes from the incremental weights.

>>>>

A second approach can be used when sampling has to be done sequentially: \textit{sequential importance resampling} (SIR). Differently from SIS, an intermediate resampling stage takes place, in which time 0 values are sampled with replacement according to a given resampling strategy (usually, sampling indexes from a multinomial distribution with probabilities equal to the weights computed through IS at time 0).\
While in the SIS all the particles (i.e., the sampled values) "survive", with the "risk" that particles associated with small weights are carried-over (in the sense that sampling in period 1 occurs from the kernel $M_1(X_0^n,dx_1)$), the same does not occur in SIR. In this second algorithm, particles with larger weights are more likely to be resampled and be used to ``generate" the particles in the next period.\
We describe the SIR more formally, focusing on the previous two-period case.
\begin{enumerate}
    \item Sample $N$ values $X_0$ from the proposal measure $\M_0$: $X_0^n\thicksim \M_0$ for $n=1,...,N$.
    \item Compute the weights $W_0^n$ for $n=1,...,N$.
    \item Sample $N$ indexes $(A_1^1,...A_1^N)$ from a multinomial distribution such that for all $j=1,...N$, $Pr(\{A_1^j=n\})=W_0^n$, for all $n=1,...N$.
    \item Build a new time 0 sample, using the values for $X_0$ associated to the previously extracted indexes $(A_1^n)_{n}$: $(X_0^{A_1^1},...,X_0^{A_1^N})$.
    \item Sample $N$ values for $X_1$ from the transition kernel $M_1$: $X_1^n\thicksim M_1(X_0^{A_1^n},dx_1)$ for $n=1,...,N$.
\end{enumerate}
Steps 3-4 constitute the actual resampling ones. At this point, let us define the new two-period sequence $\tilde X_{0:1}^n:=(X_0^{A_1^n},X_1^n)$. Then, the particle approximation for the joint distribution of $(X_0,X_1)$ becomes:
\begin{equation*}
     \Q_{1}^N(dx_{0:1})=\frac{1}{N}\sum_{n=1}^N\delta_{\tilde X_{0:1}^n}(dx_{0:1})
\end{equation*}
Note that, after resampling occurred, each particle $\tilde X_{0:1}^n:=(X_0^{A_1^n},X_1^n)$ has the same probability weight.\
To understand the logic of the previous statement, let us assume that both $\Q_0$ and $\M_0$ admit densities $q_0$, $m_0$ with respect to the same dominating measure $\nu(dx_0)$ and that the transition kernel $M_1$ also admits the density $m_1$. 
Consider the sequential updating for the weights:
\begin{equation}
    W^n_1=\frac{\Q_1(dx_{0:1})}{\M_1(dx_{0:1})}\propto \frac{q_0(x_0^{A_1^n})}{Pr(A_1^n)\cdot m_0(x_0^{A_1^n})}\cdot \frac{m_1(x_1^n|x_0^{A_1^n})}{m_1(x_1^n|x_0^{A_1^n})}
=\underbrace{\frac{q_0(x_0^{A_1^n})}{ m_0(x_0^{A_1^n})}}_{W_0^n}\cdot \underbrace{\frac{1}{Pr(A_1^n)}}_{1/W_0^n}=1\label{sis_update}
\end{equation}
where $\M_1$ denotes the probability measure for the extraction of the particle $\tilde x_{0:1}^n$ according to SIR. $\M_1$ (second equality sign in the previous formula) can be rewritten as the product between the kernel for $x_1^n$ given $x_0^{A_1^n}$ and the probability of $x_0^{A_1^n}$ itself. This last probability can also be rewritten as the product between the proposal density $m_0$ from which $X_0$ is extracted and the probability that it is sampled again when resampling occurs.

\subsection{SIS vs SIR}
The choice between the two sequential sampling strategies (SIS and SIR) has significant implications in terms of computational cost and variance of the importance weights. In fact, it is clear that SIR is more costly computationally since it requires an intermediate resampling step. Nonetheless, SIS suffers from the problem knonw as \textit{curse of dimensionality}, i.e., the fact that the $ESS$ collapses over time as the variance of the weights diverges.\
Consider an extension of SIS to a $T$-period model and let us assume that the target and the proposal kernels admit densities, respectively, $q_t$ and $m_t$ with respect to the same dominating measures (this implies that the joint distributions\footnote{Let us denote the joint distributions with $q_t$ and $m_t$ too, with some abuse of notation.} admit a density too). Then, weights, for every $t=1,...,T$, are updated according to:
\begin{equation}
    w_t^n\propto \frac{q_t(x_{0:t}^n)}{m_t(x_{0:t}^n)}\propto \underbrace{\frac{q_{t-1}( x_{0:t-1}^n)}{m_{t-1}(x_{0:t-1}^n)}}_{\propto w_{t-1}^n}\cdot \frac{q_t( x_t|x_{t-1}^n)}{m_t(x_t|x_{t-1}^n)}\propto 
    w_{t-1}^n\cdot\frac{q_t( x_t|x_{t-1}^n)}{m_t(x_t|x_{t-1}^n)}\label{sis_update}
\end{equation}
Intuitively, equation \eqref{sis_update} suggests that over time new variance is added from the incremental weights (i.e., the updating factor $q_t( x_t|x_{t-1}^n)/m_t( x_t|x_{t-1}^n)$) determining the curse of dimensionality issue. Instead, when resampling takes place, since the past weights $w_{t-1}$ are reset to be equal across them all the variance of $w_t$ inherited from the past is shut-down and the only source of variability at each stage comes from the incremental weights.
